Quantitative processing of broadband data as implemented in a scientific splitbeam echosounder

### 2023-09-27 Paper/Manuscript.tex
Updated manuscript

\usepackage{color,soul}

\usepackage{lineno}
\usepackage{xurl}

@@

\author{Lars Nonboe Andersen}
\affiliation{Kongsberg Discovery AS, Strandpromenaden 50, 3191, Horten, Norway}

\author{Dezhang Chu}
\affiliation{Fishery Resource Analysis and Monitoring Division, Northwest Fisheries Science Center, National Marine Fisheries Service, National Oceanic and Atmospheric Administration, 2725 Montlake Blvd. E. Seattle, WA, 98112, USA}

\author{Nils Olav Handegard}
\affiliation{Marine Ecosystem Acoustics, Institute of Marine Research, Bergen, 5001, Norway}

\author{Harald Heimvoll}
\affiliation{Kongsberg Discovery AS, Strandpromenaden 50, 3191, Horten, Norway}

\author{Rolf Korneliussen}
\author{Gavin J. Macaulay}
\author{Egil Ona}
\affiliation{Marine Ecosystem Acoustics, Institute of Marine Research, Bergen, 5001, Norway}

\author{Ruben Patel}
\affiliation{Codelab, Bergen, Norway}

\author{Geir Pedersen}
\email{geir.pedersen@hi.no}
\affiliation{Marine Ecosystem Acoustics, Institute of Marine Research, Bergen, 5001, Norway}

@@

Keywords:  \linebreak

@@

Echosounders are used for remote sensing marine ecosystems. As early as 1935, \citet{sund_echo_1935} observed the distribution of spawning cod in the Lofoten area using a single beam echosounder. The method was further developed to map the abundance of fish, driven by the need for fisheries management \citep{Simmonds2005Fisheries}. More recently, fisheries acoustics sensors have been deployed on a wide range of platforms, including observatories, autonomous underwater vehicles \citep{fernandes_autonomous_2003}, uncrewed surface vehicles \citep{de_robertis_uncrewed_2021} and vessels of opportunity, observing a wide range of ecosystem processes across different spatial and temporal scales \citep{godo_marine_2014}.

Today echosounders can produce pulses with a wide and continuous frequency range (broadband pulses), compared to the conventional narrowband systems. This provides in most cases significantly better along-beam resolution, a higher signal to noise ratio than narrowband pulses \citep{Chu1998Application, ehrenbergFMSlideChirp2000}, and improved frequency resolution for backscatter categorization \citep{korneliussen2018}.  \citet{lavery_measurements_2010} used broadband signals to reduce the ambiguities in interpretation of acoustic scattering from zooplankton and oceanic microstructure. \citet{Stanton2012Resonance} used low frequency broadband pulses (1-6 kHz) that included the resonance frequency of swimbladdered fish to classify size classes. \citet{blanluet_characterization_2019} used broadband acoustics, with ground truthing, to characterize the composition of two sound scattering layers (SSL). \citet{bassett_broadband_2018} showed that broadband signals may be helpful in characterizing smaller fishes with swimbladders and euphausiids. \citet{benoit-bird_exploring_2020} were able to effectively discriminate three monospecific aggregations of species (hake, anchovy and krill) using broadband signals (45-170 kHz). \citet{lavery2017} explored different broadband pulse shapes to increase the ability to resolve adjacent single targets as well as near boundaries in tank experiments. By applying high frequency broadband pulses to fish-like artificial targets, \citet{kubilius_remote_2020} demonstrated the potential for acoustic sizing of individually resolved fishes in a controlled ex situ environment. Utilizing the increased spatial resolution \citet{hasegawa_situ_2021} were able to isolate single fishes and discriminate successfully between average frequency responses of walleye pollock and pointhead flounder.

Several scientific broadband echosounder systems have been developed for laboratory use \citep{Conti2003Wide-bandwidth, Forland2014Scattering, chu1992}, some prototype or custom-made systems \citep{Zakharia1989Wide-band, Zakharia1996Wideband, Simmonds1996Species, Foote2005Measuring, Imaizumi2009Detection, Briseno-Avena2015ZOOPS, Barr2002Target} and some commercially available systems \citep{Gordon1998FishMASS, Zedel2003Acoustic, Stanton2010New, ehrenbergFMSlideChirp2000, dennyBroadbandAcousticFish1998}. More widespread use of broadband acoustics in fisheries and ecosystem research have followed from the upgrade of widely used narrowband systems to gain broadband capabilities.

Signal processing methods for broadband echosounders \citep{stanton2008} are based on radar signal processing theory, with further adaptation to echosounders \citep{lavery2017,bassett_broadband_2018}. When implementing the equations to computer processing code, some choices are well founded in the signal processing literature whereas others are of a more practical and ad-hoc nature. The latter is typically missing in the literature, making it difficult to benchmark new methods and to test the implementation in new echosounders and post processing software.

The objective of this paper and associated code is to provide a benchmark for developing and implementing signal processing algorithms for broadband echosounder. We present the design goals, implementation details, and recommended procedures and processing required to obtain quantitative broadband data. The steps include pulse compression, target strength as a function of frequency ($\ts(\freqsym)$, dB re 1m$^2$), and volume backscattering strength as a function of frequency ($\sv(\freqsym)$, dB re 1m$^-1$). The intention is that the code will be used as a starting point for implementations in various relevant data processing software, for further developing active acoustic broadband signal processing, and to serve as a learning resource.

@@

\subsection{Accompanying code}
The code  accompanying this paper is written in the Python programming language (v3.10) and is available through GitHub (see \ref{dataavstatement} for code and data availability) together with the data used in the examples. All single ping processing steps with respective figures in the paper can be reproduced by running the main script, \code{main.py}. Reproduction of Fig. \ref{fi:ts_theta_phi}a and Fig. \ref{fi:Fig_Sv_echogram}a requires downloading the original echosounder raw data separately and using the scripts (\code{TSfEchogram.py} and \code{SvfEchogram.py}). Without loss of generality, we use the \ek{} echosounder as an example since it is currently the most commonly used broadband system in the marine ecosystem acoustics field. The test data sets accompanying the code are representative for the data contained in Simrad EK80 raw files. In order to make the code echosounder independent simple json strings are used as input. Detailed information on settings used during data collection are available in the code and the data files on GitHub.

@@

\subsection{System overview}
A basic quantitative echosounder system consists of a transducer, a transceiver, and a computer program that controls the operation of the transceiver and records received signals. During transmission the program defines the signals that are created as electric signals in the transceiver, converted to acoustic signals by the transducer and transmitted into the water. The acoustic signals propagate through the water, are reflected or scattered by objects in the water, and propagate back to the transducer. During reception the transducer converts the received acoustic signals to electric signals, which are received, pre-amplified, filtered, digitized, processed in the transceiver, and then transferred to the controlling program for further data processing and storage (Fig. \ref{fi:ek_sys}). Many types of transmit signals are feasible - this paper considers only upsweep linear frequency modulated signals (also known as linear chirps).

@@

The controlling computer program generates a short-duration digital transmit signal (a ping), $\ytx(\samplesymt)$, where $\samplesymt$ is the sample index in the discrete time domain. The nominal pulse duration, $\tnom$, is defined as the duration of the digital transmit signal $\ytx(\samplesymt)$.
Typical broadband pulses are linear upsweep pulses windowed by an envelope function. The generated signal is converted to an analogue electric signal $\ytxe(\timesym)$ and amplified by the transceiver to obtain the analogue signal $\ytxe(\timesym)$, where $\timesym$ is the time for the signal. The analogue and amplified signals are passed on to the transducer to generate the transmitted acoustic signal $\ytxa(\timesym)$ in the water. For a splitbeam echosounder system, there are typically three or four channels to allow estimation of the angle of arriving echoes, and the signal is typically transmitted with equal power across the channels.

@@

The returning acoustic signal, $\yrxa(\timesym)$, is received by each transducer sector, $\channelsym$, and converted to an analog electric signal, $\yrxe(\timesym,\channelsym)$, in the transducer and received by corresponding receiver channels, $\channelsym$, in the transceiver. The received electric signal, $\yrxe(\timesym,\channelsym)$, from each channel, $\channelsym$, is pre-amplified, filtered by an analog anti-aliasing filter, and digitized in the transceiver at a frequency of $\fs$, creating the digital signal, $\yrxorg(\samplesymt,\channelsym)$.

To remove noise and reduce the quantity of data, the sampled signal from each channel is filtered and decimated in multiple stages, $\stagesym$, using complex bandpass filters, $\hbp(\genidxsym,\stagesym)$, and decimation factors, $\decfac(\stagesym)$. This processing step is approximately equivalent to a traditional demodulation process. See e.g. Table 1.2 in \citep{Demer2017} for examples of values. The individual filter coefficients for each filter and decimation stage are indexed by $\genidxsym$. The output signal from each channel, $\channelsym$, from each filter and decimation stage, $\stagesym$, is then given by:
%
\begin{equation}
\label{eq:yrx}
\yrx(\samplesymt,\channelsym,\stagesym) = \left( \yrx(\samplesymt,\channelsym,\stagesym-1) * \hbp(\genidxsym,\stagesym) \right)_{\downarrow \decfac(\stagesym)}, 
\stagesym = 1,\ldots,\nstages,
\end{equation}
%
where $\yrx(\samplesymt,\channelsym,0)$ is set to $\yrxorg(\samplesymt,\channelsym)$, being the signal before decimation, $*$ indicates convolution, $\downarrow$ indicates decimation by the factor $\decfac(\stagesym)$, and $\nstages$ is the total number of filter stages. The output signal from the final filter and decimation stage, $\yrx(\samplesymt,\channelsym,\nstages)$, is shortened to $\yrx(\samplesymt,\channelsym)$ for convenience. For the output signal, $\yrx(\samplesymt,\channelsym)$, the decimated sampling rate, $\fsdec$, is given by:
%
\begin{equation}
\label{eq:fsdec}
\fsdec = \fs\prod_{\stagesym=1}^{\nstages} \frac{1}{\decfac(\stagesym)}.
\end{equation}
%
The characteristics of the bandpass filter and decimation factors are chosen with regard to the desired operating bandwidth, noise suppression levels, impulse response duration, and other common filter characteristics, with the aim of maintaining sufficient information in the data \citep{ProakisDSP,CrochireMDSP}. The frequency responses of the filters are shown in Fig. ~\ref{fi:fir} and the corresponding filter coefficients and decimation factors are given in the test data set, where $N_v=2$.

\begin{figure}
\includegraphics[width=16cm]{Fig_fir}
\caption{\label{fi:fir} Example of frequency response (filter gain) of the filters in our test set. The blue and orange curve represents the filter response of the first and second filter. Note that the blue line is above 0 dB, this is caused by the transition from complex to real values. The vertical dashed green lines indicates the frequency range of the transmit signal.}
\end{figure}

The original sample data $\yrxorg(\samplesymt,\channelsym)$ are not available in the EK80 data files. Instead, the filtered and decimated complex samples from each transducer channel $\yrx(\samplesymt,\channelsym)$ are stored in the data files. The data are recorded in computer data files for display and analysis by processing software. Additional information, such as that from position and motion sensors and system configuration data is also included in the files.

\subsection{Matched filtering (pulse compression)}
To increase signal-to-noise ratio and resolution along the acoustic beam a matched filter may be applied to the raw data samples \citep{turin1960}. This technique is also known as pulse compression \citep{klauder1960}, achieving the average transmitted power of a relatively long pulse while obtaining the range resolution of a shorter pulse. One approach for a matched filter is to use a normalized version of the ideal transmit signal as the replicate signal, filtered and decimated using the same filters and decimation factors as applied in Eq. \ref{eq:yrx}. The normalized ideal transmit signal, $\ytxnorm(\samplesymt)$, is given by:
%
\begin{equation}
\label{eq:ytxnorm}
\ytxnorm(\samplesymt) = \frac{\ytx(\samplesymt)}{\textrm{max}(\ytx(\samplesymt))}\end{equation}
%
where $\textrm{max}$ is the maximum value of $\ytx(\samplesymt)$. The filtered and decimated output signal, $\ytxnorm(\samplesymt,\stagesym)$, from each filter stage, $\stagesym$, using the normalized ideal transmit signal, $\ytxnorm(\samplesymt)$, as the input signal, is given by:
%
\begin{equation}
\label{eq:FilterStagesTX}
\ytxnorm(\samplesymt,\stagesym) = \left[ \ytxnorm(\samplesymt,\stagesym-1) * \hbp(\genidxsym,\stagesym) \right]_{\downarrow \decfac(\stagesym)}, 
\stagesym = 1,\ldots,\nstages,
\end{equation}
%
where $\ytxnorm(\samplesymt,0)$ is set to $\ytxnorm(\samplesymt)$. The output signal from the final filter and decimation stage, $\ytxnorm(\samplesymt,\nstages)$, is used as the matched filter and is denoted as $\ymf(\samplesymt)$  (Fig.~\ref{fi:y_mf_n_ACF}a).
%
\begin{figure}
\includegraphics[width=16cm]{Fig_y_mf_n_ACF}
\caption{\label{fi:y_mf_n_ACF} An example where the decimated sampling rate ($\fsdec$, Eq. \ref{eq:fsdec}) is 125 kHz, the nominal pulse duration ($\tnom$) is 2 ms, and the effective pulse duration ($\teff$, Eq. \ref{eq:TauEff}) is 0.01 ms. (a) The absolute value of the filtered and decimated output signal, $\ymf(\samplesymt)$, from the final filter and decimation stage, that is used for the pulse compression. (b) The autocorrelation function $\ptxauto(n)$.}
\end{figure}

The auto correlation function of the matched filter signal ($\ymfauto(\samplesymt)$) and the effective pulse duration ($\teff$) will be used in later processing steps.

@@

is the square of the absolute value of the matched filter autocorrelation function, and the summation is calculated over the duration of the autocorrelation function (Fig. \ref{fi:y_mf_n_ACF}b).

To perform pulse compression the received signal, $\yrx(\samplesymt,\channelsym)$, is convolved with a complex conjugated and time-reversed version of the matched filter signal, and here also normalized with the $l^2$-norm of the matched filter to maintain received signal power. The pulse compressed signal, $\ypc(\samplesymt,\channelsym)$, then becomes
\begin{equation}
\label{eq:PulseComp}
\ypc(\samplesymt,\channelsym) = \frac{\yrx(\samplesymt,\channelsym)*\ymf^*(-\samplesymt)}{||\ymf||^2_2},
\end{equation}
%
The received power samples are then used to estimate target strength and volume backscattering strength. For estimating received power samples, the average signal, $\ypc(\samplesymt)$, over all transducer sectors, $\nchannels$, is used:

@@

Compensation of echo strength for position in the acoustic beam requires an estimate of the echo arrival angle. This is obtained using the splitbeam method \citep{burdic1991}, which for broadband pules can be implemented with the angle values contained in the complex-valued $\ypc(\samplesymt)$ data, in combination with knowledge of transducer sector geometry. The principle is demonstrated with a transducer that is divided into four quadrants (Fig. \ref{fi:Fig_trd_quad_Impedances}a). In this example the summed signals from four halves (1+2, 2+3, 3+4, 4+1) are calculated as:
\begin{eqnarray}
\label{eq:SumHalves}
y_{\textrm{pc,fore}}(\samplesymt) & = & \frac{1}{2} \left( \ypc(\samplesymt,3)+\ypc(\samplesymt,4) \right),\\
y_{\textrm{pc,aft}}(\samplesymt) & = & \frac{1}{2} \left( \ypc(\samplesymt,1)+\ypc(\samplesymt,2) \right),\\
y_{\textrm{pc,star}}(\samplesymt) & = & \frac{1}{2} \left( \ypc(\samplesymt,1)+\ypc(\samplesymt,4) \right),\\
y_{\textrm{pc,port}}(\samplesymt) & = & \frac{1}{2} \left( \ypc(\samplesymt,2)+\ypc(\samplesymt,3) \right),
\end{eqnarray}
%
where fore, aft, star(board), and port indicate the relevant transducer halves.

\begin{figure}
\includegraphics[width=16cm]{Fig_trd_quad_Impedances}
\caption{\label{fi:Fig_trd_quad_Impedances}(a) Transducer divided into four quadrants. The labels are directions often used when a transducer is mounted on a ship. (b) Equivalent circuit diagram of transducer/transceiver with system impedances.}
\end{figure}
%

\subsection{Power and angle samples}
The transceiver measures voltage over a load, $\zrxe$, connected in series with the transducer impedance, $\ztde$. When calculating various acoustic properties, a system gain parameter will be used which assumes a matched receiver load. The total received power, $\prxe(\samplesymt)$, from all transducer sectors for a matched receiver load (Fig. \ref{fi:Fig_trd_quad_Impedances}b) is given by: 
\begin{equation}
\label{eq:prx}
\prxe(\samplesymt) = \nchannels\left( \frac{|\ypc(\samplesymt)|}{2\sqrt{2}} \right)^2 \left( \frac{|\zrxe+\ztde|}{\zrxe} \right)^2 \frac{1}{|\ztde|}.
\end{equation}
%

>>>>

\usepackage{color,soul}
\usepackage{lineno}
\usepackage{xurl}
\usepackage{hyperref}

@@

@@

@@

Echosounders are used for remote sensing marine ecosystems. As early as 1935, \citet{sund_echo_1935} observed the distribution of spawning cod in the Lofoten area using a single beam echosounder. The method was further developed to map the abundance of fish, driven by the need for fisheries management \citep{Simmonds2005Fisheries}. More recently, fisheries acoustics sensors have been deployed on a wide range of platforms including observatories, autonomous underwater vehicles \citep{fernandes_autonomous_2003}, uncrewed surface vehicles \citep{de_robertis_uncrewed_2021} and vessels of opportunity, observing a wide range of ecosystem processes across different spatial and temporal scales \citep{godo_marine_2014}.

Today, echosounders can produce pulses with a wide and continuous frequency range (broadband pulses), compared to the conventional narrowband systems. This provides in most cases significantly better along-beam (range) resolution, a higher signal-to-noise ratio than narrowband pulses \citep{Chu1998Application, ehrenbergFMSlideChirp2000}, and improved frequency resolution for backscatter categorization \citep{korneliussen2018}.  \citet{lavery_measurements_2010} used broadband signals to reduce ambiguities in the interpretation of acoustic scattering from zooplankton and oceanic microstructure. \citet{Stanton2012Resonance} used low frequency broadband pulses (1-6 kHz) that included the resonance frequency of swimbladdered fish to classify size classes. \citet{blanluet_characterization_2019} used broadband acoustics, with ground truthing (nets and video), to characterize the composition of two sound scattering layers (SSL). \citet{bassett_broadband_2018} showed that broadband signals may be helpful in characterizing smaller fishes with swimbladders and euphausiids (15-150 kHz). \citet{benoit-bird_exploring_2020} were able to effectively discriminate three monospecific aggregations of species (hake, anchovy and krill) using broadband signals (45-170 kHz). \citet{lavery2017} explored different broadband pulse shapes to increase the ability to resolve adjacent single targets as well as near boundaries in tank experiments (15-400 kHz). By applying high-frequency broadband pulses to fish-like artificial targets, \citet{kubilius_remote_2020, kubilius_remote_2023} demonstrated the potential for acoustic sizing of individually resolved fishes in a controlled \textit{ex situ} environment (45-90 and 160-260 kHz). Using the increased range resolution, \citet{hasegawa_situ_2021} were able to isolate single fishes and discriminate successfully between average frequency responses of walleye pollock and pointhead flounder \textit{in situ} (45-260 kHz). Using narrow (18, 38 kHz) and broadband acoustic data (70-280 kHz) with a mixed species scattering model,  \citet{loranger_broadband_2022} demonstrated estimation of mean length and total biomass of longfin squid and mackerel.

Several scientific broadband echosounder systems have been developed for laboratory use \citep{Conti2003Wide-bandwidth, Forland2014Scattering, chu1992}, some prototype or custom-made systems \citep{Zakharia1989Wide-band, Zakharia1996Wideband, Simmonds1996Species, Foote2005Measuring, Imaizumi2009Detection, Briseno-Avena2015ZOOPS, Barr2002Target} and some commercially available systems \citep{Gordon1998FishMASS, Zedel2003Acoustic, Stanton2010New, ehrenbergFMSlideChirp2000, dennyBroadbandAcousticFish1998}. More widespread use of broadband acoustics in fisheries and ecosystem research has followed from the upgrade of widely used narrowband systems to gain broadband capabilities.

Signal processing methods for broadband echosounders \citep{stanton2008} are based on radar signal processing theory, with further adaptation to echosounders \citep{lavery2017,bassett_broadband_2018}. When implementing the equations to computer processing code, some choices are well founded in the signal processing literature, whereas others are of a more practical and ad-hoc nature. The latter is typically missing in the literature, making it difficult to benchmark new methods and to test the implementation in new echosounders and post-processing software.

The objective of this paper and the associated code is to provide a benchmark for developing and implementing signal processing algorithms for broadband echosounders. We present the design goals, implementation details, and recommended procedures and processing required to obtain quantitative broadband data. The steps include pulse compression, target strength as a function of frequency ($\ts(\freqsym)$, dB re 1m$^2$), and volume backscattering strength as a function of frequency ($\sv(\freqsym)$, dB re 1m$^{-1}$). The intention is that the code will be used as a starting point for implementations in various relevant data processing software, for further developing active acoustic broadband signal processing, and to serve as a learning resource.

@@

\subsection{Accompanying code}
The code  accompanying this paper is written in the Python programming language (v3.10) and is available through GitHub (see \autoref{dataavstatement} for code and data availability) together with the data used in the examples. All single-ping processing steps with respective figures in the paper can be reproduced by running the main script, \code{main.py}. Reproduction of \autoref{fi:ts_theta_phi}a and \autoref{fi:Fig_Sv_echogram}a requires downloading the original echosounder raw data separately and using the scripts (\code{TSfEchogram.py} and \code{SvfEchogram.py}).

Without loss of generality, we use the \ek{} echosounder as an example, since it is currently the most commonly used broadband system in the marine ecosystem acoustics field. The test data sets accompanying the code are representative of the data contained in Simrad EK80 raw files. In order to make the code echosounder independent, simple json strings are used as input. Detailed information on the settings used during data collection is available in the code and data files on GitHub.

@@

\subsection{System overview}
A basic quantitative echosounder system consists of a transducer, a transceiver, and a computer program that controls the operation of the transceiver and records the received signals. During transmission, the program defines the signals that are created as electric signals in the transceiver, converted to acoustic signals by the transducer and transmitted into the water. The acoustic signals propagate through the water, are reflected or scattered by objects in the water, and propagate back to the transducer. During reception, the transducer converts the received acoustic signals to electric signals, which are received, preamplified, filtered, digitized, processed in the transceiver, and then transferred to the controlling program for further data processing and storage ( \autoref{fi:ek_sys}). Many types of transmit signals are feasible - this paper considers only upsweep linear frequency modulated signals (also known as linear chirps).

@@

The controlling computer program generates a short-duration digital transmit signal (a ping), $\ytx(\samplesymt)$, where $\samplesymt$ is the sample index in the discrete time domain. The nominal pulse duration, $\tnom$, is defined as the duration of the digital transmit signal $\ytx(\samplesymt)$.
Typical broadband pulses are linear upsweep pulses windowed by an envelope function. The generated signal is converted to an analogue electric signal $\ytxe(\timesym)$ and amplified by the transceiver to obtain the analogue signal $\ytxe(\timesym)$, where $\timesym$ is the time for the signal. The analogue and amplified signals are passed on to the transducer to generate the transmitted acoustic signal $\ytxa(\timesym)$ in the water. For a split-beam echosounder system, there are typically three or four channels to allow estimation of the angle of arriving echoes, and the signal is typically transmitted with equal power across the channels.

@@

The returning acoustic signal, $\yrxa(\timesym)$, is received by each transducer sector, $\channelsym$, and converted to an analog electric signal, $\yrxe(\timesym,\channelsym)$, in the transducer and received by the corresponding receiver channels, $\channelsym$, in the transceiver. The received electric signal, $\yrxe(\timesym,\channelsym)$, from each channel, $\channelsym$, is pre-amplified, filtered by an analog anti-aliasing filter, and digitized in the transceiver at a frequency of $\fs$, creating the digital signal, $\yrxorg(\samplesymt,\channelsym)$.

To remove noise and reduce the amount of data, the sampled signal from each channel is filtered and decimated in multiple stages, $\stagesym$, using complex bandpass filters, $\hbp(\genidxsym,\stagesym)$, and decimation factors, $\decfac(\stagesym)$. This processing step is approximately equivalent to a traditional demodulation process. See e.g. Table 1.2 in \citep{Demer2017} for examples of values. The individual filter coefficients for each filter and decimation stage are indexed by $\genidxsym$. The output signal from each channel, $\channelsym$, from each filter and decimation stage, $\stagesym$, is then given by:
%
\begin{equation}
\label{eq:yrx}
\yrx(\samplesymt,\channelsym,\stagesym) = \left( \yrx(\samplesymt,\channelsym,\stagesym-1) * \hbp(\genidxsym,\stagesym) \right)_{\downarrow \decfac(\stagesym)}, 
\stagesym = 1,\ldots,\nstages,
\end{equation}
%
where $\yrx(\samplesymt,\channelsym,0)$ is set to $\yrxorg(\samplesymt,\channelsym)$, being the signal before decimation, $*$ indicates convolution, $\downarrow$ indicates decimation by the factor $\decfac(\stagesym)$, and $\nstages$ is the total number of filter stages. The output signal from the final filter and decimation stage, $\yrx(\samplesymt,\channelsym,\nstages)$, is shortened to $\yrx(\samplesymt,\channelsym)$ for convenience. For the output signal, $\yrx(\samplesymt,\channelsym)$, the decimated sampling rate, $\fsdec$, is given by:
%
\begin{equation}
\label{eq:fsdec}
\fsdec = \fs\prod_{\stagesym=1}^{\nstages} \frac{1}{\decfac(\stagesym)}.
\end{equation}
%
The characteristics of the bandpass filter and decimation factors are chosen with regard to the desired operating bandwidth, noise suppression levels, impulse response duration, and other common filter characteristics, with the aim of maintaining sufficient information in the data \citep{ProakisDSP,CrochireMDSP}. The frequency responses of the filters are shown in \autoref{fi:fir} and the corresponding filter coefficients and decimation factors are given in the test data set, where $N_v=2$.

\begin{figure}
\includegraphics[width=16cm]{Fig_fir}
\caption{\label{fi:fir} Example of frequency response (filter gain) of the filters in our test set. The blue and orange curves represent the filter responses of the first and second filter. Note that the blue line is above 0 dB; this is caused by the transition from complex to real values. The vertical dashed green lines indicate the frequency range of the transmit signal.}
\end{figure}

The original sample data $\yrxorg(\samplesymt,\channelsym)$ are not available in the EK80 data files. Instead, the filtered and decimated complex samples from each transducer channel $\yrx(\samplesymt,\channelsym)$ are stored in the data files. Data are recorded in computer data files for display and analysis by processing software. Additional information, such as from position and motion sensors and system configuration data, is also included in the files.

\subsection{Matched filtering (pulse compression)}
To increase signal-to-noise ratio and resolution along the acoustic beam, a matched filter may be applied to the raw data samples \citep{turin1960}. This technique is also known as pulse compression \citep{klauder1960}, which achieves the average transmitted power of a relatively long pulse while obtaining the range resolution of a shorter pulse. One approach for a matched filter is to use a normalized version of the ideal transmit signal as the replicate signal, filtered and decimated using the same filters and decimation factors as applied in \autoref{eq:yrx}. The normalized ideal transmit signal, $\ytxnorm(\samplesymt)$, is given by:
%
\begin{equation}
\label{eq:ytxnorm}
\ytxnorm(\samplesymt) = \frac{\ytx(\samplesymt)}{\textrm{max}(\ytx(\samplesymt))}\end{equation}
%
where $\textrm{max}$ is the maximum value of $\ytx(\samplesymt)$. The filtered and decimated output signal, $\ytxnorm(\samplesymt,\stagesym)$, of each filter stage, $\stagesym$, using the ideal normalized transmit signal, $\ytxnorm(\samplesymt)$, as the input signal, is given by:
%
\begin{equation}
\label{eq:FilterStagesTX}
\ytxnorm(\samplesymt,\stagesym) = \left[ \ytxnorm(\samplesymt,\stagesym-1) * \hbp(\genidxsym,\stagesym) \right]_{\downarrow \decfac(\stagesym)}, 
\stagesym = 1,\ldots,\nstages,
\end{equation}
%
where $\ytxnorm(\samplesymt,0)$ is set to $\ytxnorm(\samplesymt)$. The output signal from the final filter and decimation stage, $\ytxnorm(\samplesymt,\nstages)$, is used as the matched filter and is indicated as $\ymf(\samplesymt)$  (\autoref{fi:y_mf_n_ACF}a).
%
\begin{figure}
\includegraphics[width=16cm]{Fig_y_mf_n_ACF}
\caption{\label{fi:y_mf_n_ACF} An example where the decimated sampling rate ($\fsdec$, \autoref{eq:fsdec}) is 125 kHz, the nominal pulse duration ($\tnom$) is 2 ms, and the effective pulse duration ($\teff$, \autoref{eq:TauEff}) is 0.01 ms. (a) The absolute value of the filtered and decimated output signal, $\ymf(\samplesymt)$, from the final filter and decimation stage, which is used for the pulse compression. (b) The autocorrelation function $\ptxauto(n)$.}
\end{figure}

The autocorrelation function of the matched filter signal ($\ymfauto(\samplesymt)$) and the effective pulse duration ($\teff$) will be used in later processing steps.

@@

is the square of the absolute value of the matched filter autocorrelation function, and the summation is calculated over the duration of the autocorrelation function (\autoref{fi:y_mf_n_ACF}b).

To perform pulse compression, the received signal, $\yrx(\samplesymt,\channelsym)$, is convolved with a complex conjugated and time-reversed version of the matched filter signal, and here also normalized with the $l^2$-norm of the matched filter to maintain received signal power. The pulse compressed signal, $\ypc(\samplesymt,\channelsym)$, then becomes
\begin{equation}
\label{eq:PulseComp}
\ypc(\samplesymt,\channelsym) = \frac{\yrx(\samplesymt,\channelsym)*\ymf^*(-\samplesymt)}{||\ymf||^2_2},
\end{equation}
%
The received power samples are then used to estimate target strength and volume backscattering strength. For estimating received power samples, the average signal, $\ypc(\samplesymt)$, over all transducer sectors, $\nchannels$, is used:

@@

Compensation of echo strength for position in the acoustic beam requires an estimate of the echo arrival angle. This is obtained using the split-beam method \citep{burdic1991}, which for broadband pules can be implemented with the angle values contained in the complex-valued $\ypc(\samplesymt)$ data, in combination with knowledge of transducer sector geometry. The principle is demonstrated with a transducer divided into four quadrants (\autoref{fi:Fig_trd_quad_Impedances}a). In this example, the summed signals from four halves (1+2, 2+3, 3+4, 4+1) are calculated as:
\begin{eqnarray}
\label{eq:SumHalves}
y_{\textrm{pc,fore}}(\samplesymt) & = & \frac{1}{2} \left( \ypc(\samplesymt,3)+\ypc(\samplesymt,4) \right),\\
y_{\textrm{pc,aft}}(\samplesymt) & = & \frac{1}{2} \left( \ypc(\samplesymt,1)+\ypc(\samplesymt,2) \right),\\
y_{\textrm{pc,star}}(\samplesymt) & = & \frac{1}{2} \left( \ypc(\samplesymt,1)+\ypc(\samplesymt,4) \right),\\
y_{\textrm{pc,port}}(\samplesymt) & = & \frac{1}{2} \left( \ypc(\samplesymt,2)+\ypc(\samplesymt,3) \right),
\end{eqnarray}
%
where fore, aft, star(board), and port indicate the relevant transducer halves. In scientific echoshounder data, fore-aft angles are often labelled  alongship angles and  port-starboard angles athwartship angles.

\begin{figure}
\includegraphics[width=16cm]{Fig_trd_quad_Impedances}
\caption{\label{fi:Fig_trd_quad_Impedances}(a) Transducer divided into four quadrants. The labels are directions often used when a transducer is mounted on a ship. (b) Equivalent circuit diagram of the transducer/transceiver with the impedances of the system.}
\end{figure}
%

\subsection{Power and angle samples}
The transceiver measures voltage over a load, $\zrxe$, connected in series with the transducer impedance, $\ztde$. When calculating various acoustic properties, a system gain parameter will be used that assumes a matched receiver load. The total received power, $\prxe(\samplesymt)$, from all transducer sectors for a matched receiver load (\autoref{fi:Fig_trd_quad_Impedances}b) is given by: 
\begin{equation}
\label{eq:prx}
\prxe(\samplesymt) = \nchannels\left( \frac{|\ypc(\samplesymt)|}{2\sqrt{2}} \right)^2 \left( \frac{|\zrxe+\ztde|}{\zrxe} \right)^2 \frac{1}{|\ztde|}.
\end{equation}
%

Forward/aft and port/starboard phase angles of target echoes are estimated by combining the transducer half signals thus: 
%
\begin{eqnarray}
\label{eq:phase1}
y_\along(\samplesymt) & = & y_{\textrm{pc,fore}}(\samplesymt) y_{\textrm{pc,aft}}^*(\samplesymt), \\
y_\athw(\samplesymt) & = & y_{\textrm{pc,star}}(\samplesymt) y_{\textrm{pc,port}}^*(\samplesymt),
\end{eqnarray}
%
where $y_\along(\samplesymt)$ is the electrical angle along the minor axis of the transducer (positive in the forward direction when ship-mounted) and $y_\athw(\samplesymt)$ the electrical angle along the major axis of the transducer (positive to starboard when ship-mounted). The physical echo arrival angles ($\along$ and $\athw$) are then given by:
%
\begin{eqnarray}
\label{eq:phase2}
\along(\samplesymt) & = & \arcsin\left( \frac{\atan\left( \Im(y_\along(\samplesymt)), \Re(y_\along(\samplesymt) \right)}{\anglefalong}\right) \\
\athw(\samplesymt) & = & \arcsin\left( \frac{\atan\left( \Im(y_\athw(\samplesymt)), \Re(y_\athw(\samplesymt) \right)}{\anglefathw}\right),
\end{eqnarray}
%
where $\anglefalong$ and $\anglefathw$ are constants that convert from phase angles to physical echo arrival angles (\autoref{fi:ts_theta_phi}b) and are derived from the geometry of the transducer \citep{Urick3rdPrinciples} and $\fc$ the centre frequency of the chirp pulse \citep{ehrenberg1979}. The inverse sine is indicated by $\arcsin$,the four quadrant inverse tangent which returns values in the interval $[-\pi, \pi]$ inclusive is indicated by $\atan$, the real part of a complex number by $\Re$ and the imaginary part by $\Im$. As a mnemonic, the horizontal line in the symbol used for the forward/aft direction, $\along$, represents the pivot axis for the alongship angles and the near-vertical line in the $\athw$ symbol indicates the pivot axis for the port/starboard angles.
%
